---
title: "GGPLOT Best Practices"
output: html_notebook
---

```{r}
library(ggplot2)
str(mtcars)
install.packages('Hmisc', repos = "http://cran.us.r-project.org")
```


## Barplots

We want to avoid _Dynamite Plots_ as shown here:
```{r}
# Base layers
m <- ggplot(mtcars, aes(x = cyl, y = wt))

# Draw dynamite plot
m +
  stat_summary(fun.y = mean, geom = 'bar', fill = 'skyblue') +
  stat_summary(fun.data = mean_sdl, fun.args = list(mult = 1), geom = "errorbar", width = 0.1)
```
These plots could give the impression that there are data points at 1 or 0. Of course there are no cars that weigh 1kg. More importantly the bars do not show data above 4 in this case although the error bars indicate higher max values (the max is 5.242). Finally, we cannot see how much datapoints are there for each cyliner based on the plot.


## Scatterplots
A better alternative to get the same and even more information is plotting the individual data points. Note that the cylinders are 4,6 and 8 and that the points are spread to avoid _overplotting_. This is a good alternative for relatively small data sets and variables that are normally distributed, not having extreme outliers. 

```{r}
ggplot(mtcars, aes(x=cyl,y=wt)) +
  geom_jitter(width=0.2) +
  stat_summary(fun.y = mean, geom = 'point', fill = 'red') +
  stat_summary(fun.data = mean_sdl, fun.args = list(mult = 1), geom = "errorbar", width = 0.2, col='red')
```

## Boxplots
One of the best methods of gaining an overview of the data is using boxplots. There is a disadvantage of the standard boxplot, namely that we again cannot see the size of n. We can circumvent this disadvantage by applying the following boxplots, because the width of the plot reflects the size of n:

```{r}
ggplot(mtcars,aes(x=cyl,y=wt, group=cyl,))+
  geom_boxplot(varwidth = T, fill='skyblue')
```

Other plots that I might explain later.

## Density and violin plot
```{r}
# Unweighted density plot from before
ggplot(msleep, aes(x = sleep_total, fill = vore)) +
  geom_density(col = NA, alpha = 0.35) +
  scale_x_continuous(limits = c(0, 24)) +
  coord_cartesian(ylim = c(0, 0.3))

# Unweighted violin plot
ggplot(msleep, aes(x = vore, y = sleep_total, fill = vore)) +
  geom_violin()

# Calculate weighting measure
library(dplyr)
mammals2 <- msleep %>%
  group_by(vore) %>%
  mutate(n = n() / nrow(msleep)) -> msleep

# Weighted density plot
ggplot(mammals2, aes(x = sleep_total, fill = vore)) +
  geom_density(aes(weight = n), col = NA, alpha = 0.35) +
  scale_x_continuous(limits = c(0, 24)) +
  coord_cartesian(ylim = c(0, 0.3))

# Weighted violin plot
ggplot(mammals2, aes(x = vore, y = sleep_total, fill = vore)) +
  geom_violin(aes(weight = n), col = NA)
```

## 2D density plots
```{r}
# Base layers
p <- ggplot(faithful, aes(x = waiting, y = eruptions)) +
  scale_y_continuous(limits = c(1, 5.5), expand = c(0, 0)) +
  scale_x_continuous(limits = c(40, 100), expand = c(0, 0)) +
  coord_fixed(60 / 4.5)

# 1 - Use geom_density_2d()
p + geom_density_2d()

# 2 - Use stat_density_2d() with arguments
p + stat_density_2d(aes(col = ..level..), h = c(5, 0.5))
```

```{r}
# Load in the viridis package
library(viridis)

# Add viridis color scale
ggplot(faithful, aes(x = waiting, y = eruptions)) +
  scale_y_continuous(limits = c(1, 5.5), expand = c(0,0)) +
  scale_x_continuous(limits = c(40, 100), expand = c(0,0)) +
  coord_fixed(60/4.5) +
  stat_density_2d(geom = "tile", aes(fill = ..density..), h=c(5,.5), contour = FALSE) +
  scale_fill_viridis()
```

